Detecting skill level and mental template in biface morphology: Archaeological and experimental insights

### 2022-06-06 manuscript/manuscript.Rmd
finished the very first draft.

---
title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Stone tools provide key evidence of human cognitive evolution but remain challenging to interpret. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute the great variation of biface morphology, encompassing raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

Skill level: short literature review. Symmetry, thinning,

Mental template: short literature review

Here we have two interconnected research questions based on a reference sample generated from our 90-hour handaxe skill acquisition experiment: 1) Can skill level and mental template (or "aesthetic preference") be effeciently detected from biface morphometric data? 2) What is the effect of training on these two aspects?

@@

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through Neolithics.com into flat blanks of similar size and shape for training and assessments.

@@

PC1 (thickness) is a robust indicator of skill level as it is shared by modern expert knapper and boxgrove foragers, while PC2 (elongation and pointedness) reflects more of personal/community-level aesthetic choices. However, we do not intend to construct a false binary framework and put these two factors as disconnected and opposite concepts.

In terms of our second research question, this study shows that training do have an immediate intervention effects (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nontheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate peadagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by Late Acheulean toolmaker as a new benchmark to deepen our understanding on this issue. It is noteworthy that how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group. Some potential explainations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].

The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to a desired form (**Figure** \@ref(fig:comparison)). That being said, if the given nodules already possess a oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group display the lowest value (**Figure** \@ref(fig:deltaweight)). On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by foragers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

Experimental archaeology has a huge potential in similar topics.

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R. Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas. Given the irreversible nature of archaeological excavation, the digitized data, be it text, pictures, or videos, often become the sole evidence that are available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons of preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue through sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data resue is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to more detailed technological analysis of stone artifacts, however, our paper shows that finding the approriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contempoarary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an effect, but 90 hours of training is still not enough for novice to reach the level of expertise.

>>>>

---
title: "Detecting skill level and mental template in biface morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Radu Iovita^[Department of Anthropology, New York University, New York, NY, USA; Department of Early Prehistory and Quaternary Ecology, Eberhard Karls University of Tübingen, Tübingen, Germany; iovita@nyu.edu]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Palaeo-Research Institute, University of Johannesburg, Auckland Park, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean bifaces, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Among many factors identified to contribute to the considerable variation of biface morphology, skill level and mental template have been frequently quoted and discussed. In this paper, we compared the experimental replicas made by modern expert and novice knappers with the Boxgrove handaxe assemblage. Through the multivariate analysis of morphometric data, our study suggested that both skill level and mental template have a relatively clear manifestation in different aspects of biface morphology, where the former is related to thickness while the latter is connected with elongation and pointedness. Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and found that reaching the level of modern expert requires more training time than what is permitted in the experimental setting. Our study demonstrated the potential of experimental archaeology and digitized secondary data in revealing new insights from old archaeological assemblages. \\par \\textbf{Keywords:} Late Acheulean; Biface production; Boxgrove; Experimental archaeology; Skill level; Mental template"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean bifaces has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the heterogeneity and longevity of biface assemblage on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to the great variation of biface morphology, including but not limited to raw materials [@sharon2008; @eren2014; @lycett2016], hammer stone [@shipton2009], function [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], skill level [@stout2014; @herzlinger2017; @caruana2021], mental template [@garcía-medrano2019; @hutchence2021], among which skill level and mental template have been repeatedly mentioned and discussed in the now extensive corpus of biface studies.

In its classical definition, mental template is the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept essentially lies in the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorse this conceptual framework and actively applies in the typological analysis of biface at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental template (or "imposed form") has been recognized as two major technological innovations of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for its normative and static assumption [@lyman2004], ignorance of other competing factors such as raw material constraint [@white1995], and the lacking of rigorous studies of its corresponding cognitive processes. Until very recently, several researchers have actively addressed these critiques and reconceptualized the mental template in the study of biface morphology. Hutchence and Scott [-@hutchence2021: 675], for example, leveraged the theory of "community of practice" [@wenger1998] and argued that "the consistency in handaxe shape found at sites like Boxgrove is a consequence of the emergent actions of individual knappers being simultaneously constrained by the imposition of social norms." Furthermore, raw material is often treated as a crucial variable to be tested at the very beginning of a research design focusing on mental template [@garcía-medrano2019]. In terms of the cognitive mechanisms behind mental template, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations, which has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors. In short, when exercised with proper caution, mental template still has its value in our study of biface morphological variation.

Building upon the concept of mental template, one possible way of defining skill is the capacity of realizing the former using the resources available [@roux1995: 66]. This version of conceptualization, particularly relevant when it comes to motor skills such as knapping, can be dismantled into two mutually dependent aspects, namely the intentional aspect (goal/strategic planning) and the operational aspect (means/motor execution) [@connolly1989]. It also roughly corresponds to the well-known dichotomy developed by French lithic analysts of "*connaissance*" (abstract knowledge) and "*savoir-faire*" (practical know-how) [@pelegrin1993]. As Stout [-@stout2002: 694] noted, the acquisition of skill is deeply rooted in its social context, and it is not composed of "some rigid motor formula" but "how to act in order to solve a problem". This ecological notion of skill somewhat mirrors Hutchence and Scott's [-@hutchence2021] reconceptualization of mental template in that they both refute the idea that technology is simply an internal program expressed by mind and prefer a dynamic approach emphasizing the interaction between the perception and action. The manifestations of skill in materialized form display a great amount of variation, but ethnoarchaeological studies have repeatedly suggested that skills can be improved through practice as perceived by the local practitioners and it is thus possible to evaluate the skill level reflected in the knapping products based on time spent on practice [@stout2002; @roux1995]. When contextual information is less readily available as in many technological systems, how to properly operationalize and measure knapping skills has been a methodological issue receiving much attention among archaeologists [@bamforth2008; @kolhatkar2022]. In the context of biface technology, in addition to measurements that can be almost applied in any lithic technological system such as raw material and hammer choices, platform preparation, as well as hinges, symmetry [@hodgson2015; @hutchence2019] and thinning [@caruana2020; @stout2014] have been frequently quoted as reliable and distinctive indicators of the skill level as supported very several experimental studies. These two have also been commonly used as standards for dividing Early Acheulean and Late Acheulean [@schick1993; @clark2001; @callahan1979].

Drawing on these two lines of literature, we aim to explore the possibility of differentiating skill level and mental template through a comparative study of an archaeological biface assemblage known for its remarkable dexterity, a reference biface collection produced by modern knapping experts, and an experimental biface sample produced by modern novice knappers. Since the novice biface collection is generated from a 90-hour skill acquisition experiment, we also have the precious opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental template be efficiently detected from biface morphometric data? 2) How does training affect novices' performance in these two aspects?

@@

The biface experimental replicas used in this study comprised two sub-collection. The first sub-collection includes 10 bifaces knapped by three expert knappers, including Bruce Bradley (n=4), John Lord (n=3), and Dietrich Stout (n=3) [@stout2014]. The second sub-collection is produced from a 90-hour handaxe knapping skill acquisition experiment [@bayani2021; @pargeter2019; @pargeter2020], where 30 adults with no previous experience in knapping were recruited from Emory University and its surrounding communities and requested to make 132 bifaces in total. Among these 30 adult participants, 17 have gone through multiple one-to-one or group training sessions that amounted to 89 hours in maximum, while the remaining 13 were assigned to the controlled group, where no formal training is given. As part of the preparation efforts, the experimental team spalled the Norfolk flints acquired through [Neolithics.com](https://neolithics.com/) into flat blanks of similar size and shape for training and assessments. The mechanical properties of these raw materials are comparable to the ones used in Boxgrove in that they are both fine-grained and highly predictable in fracturing process.

@@

In accordiance with the existing literature on biface knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (thickness) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove foragers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the biface into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of biface produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394]. This also reflects the majority of samples in both our expert and novice experimental assemblages. In the skill acquisition experiments, novice knappers have been explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], corresponding to the cases of replicas produced by Bruce Bradley and a few novices in our study.

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. This divergence of group-level aesthetic choices can be best explained under the theoretical framework of the communities of practice [@wenger1998] as advocated by Hutchence and Scott in biface analysis[-@hutchence2021]. The most common form of learning in the experiment occurred in the group condition, where the instructor taught multiple novices at the same time and novices have the chance to also communicated and learned from their peers. Unfortunately, the biface data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather unconsipicous. This finding corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. This follow-up project further adds the samples produced by the Late Acheulean toolmaker as a new benchmark to deepen our understanding of this issue. It is noteworthy how constrained is the Boxgrove assemblage morphological variation as measured by both PC1 and PC2 even when compared with the modern expert group (**Figure** \@ref(fig:GeneralPCA1)), especially given the fact that it has the largest sample size among all studied group. Some potential explanations for this phenomenon include 1) the strong idiosyncrasy of individual expert knappers shaped by their own unique learning and practice experience; and/or 2) the present day-skill shortage of our expert knapper as compared with Boxgrove knappers despite their multiple years of knapping practice [@milks2019].

The pre-training group is similar to the Boxgrove group in PC2 because these novices lack the ability to effectively reduce the nodules, which are typically flat pre-prepared cortical flakes, to the desired form (**Figure** \@ref(fig:comparison)). That being said, if the given nodules already possess an oval morphology like those presented in the Boxgrove assemblage, it is likely the form of end products knapped by novices in the pre-training group will remain roughly unchanged. This explanation is also supported by the comparison of average delta weight, defined as the difference between the weight of handaxe and the weight of nodule, among four groups, where the pre-training group displays the lowest value (**Figure** \@ref(fig:deltaweight)). On the other hand, the refitting analyses of the Boxgrove handaxe assemblage have suggested that the nodules exploited by foragers inhabiting this site are somewhat bulky and amorphous [@roberts1998: 339, 360]. These characteristics have been clearly displayed in a recent attempt of slow-motion refitting of a handaxe specimen from Boxgrove GTP17 (<https://www.youtube.com/watch?v=iS58MUJ1ZEo>). As such, behind the resemblance of the pre-training group and the Boxgrove assemblage in PC2 are two types of mechanisms that are fundamentally different from each other, where the latter group exhibits a complex suite of cognitive and motor execution processes to transform the shapeless raw materials to a delicate end product in a given shape.

@@

Another contribution that we would like to highlight here is that this research project demonstrates the potential of reusing old archaeological data in digital format to address novel research questions. In this paper, the main source of archaeological data is a collection of photos produced and curated by one of our co-authors (R. Iovita) more than 10 years ago, and the morphological variation data of the experimental collection are also derived from photographs instead of remeasurements of the original replicas. Given the irreversible nature of archaeological excavations, digitized data, be it text, pictures, or videos, often become the sole evidence that is available for certain research questions. Yet, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018]. As stated in the method section, we addressed this issue by sharing the raw data and the code for generating the derived data on an open-access repository. Another major and legitimate concern of archaeological data reuse is their quality. In terms of this aspect, we do acknowledge the limitations of relying on photos when it comes to the more detailed technological analysis of stone artifacts, however, our paper shows that finding the appropriate research questions given the data available is key to revealing new novel insights into the studied topic. Moreover, we believe that this type of research has a strong contemporary relevance due to the continued influence of the COVID-19 on fieldwork-related travel and direct access to archaeological artifacts [@balandier2022; @ogundiran2021].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with thickness while the latter is reflected in elongation and pointedness; 2) Training has an immediate effect of sharing a common mental template, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove foragers. However, it should be noted that it is not our intention to construct a false binary framework and put these two factors as disconnected and opposite concepts. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of bifaces as well as their implications for the biological and cultural evolution of the hominin lineages.
